---
title: "Experimental Design and Data Analysis 2021/22"
subtitle: "(02) R Code Basics"
author: "Axel Nohturfft"
date: "`r Sys.Date()`"
output: 
  html_notebook:
    toc: TRUE
    toc_float: TRUE
    toc_depth: 3
    code_folding: "show"
---

## Topics covered in this script  

* Using scripts  
* Scalars and vectors  
* Functions  
* Variables  


```{css echo=FALSE}
h1 {background: brown;color: white;padding-left: 7px;}
h2 {background: grey;color: white;padding-left: 7px;}
h3 {color: brown;}
code {color: chocolate;}
```

```{r}
# Clear variables in memory
rm(list = ls(all=TRUE))
```


## Scripts  
***ALWAYS SAVE YOUR CODE IN SCRIPTS !***  

* To document your work (think of them as a data analyst's lab notebook)
* To quickly repeat an analysis in the future
* To re-use your code for new analyses
* To generate reports (rmarkdown scripts)

There are two types of R scripts:  

* Standard R scripts (see `r_code_basics.R`)  
* rmarkdown (`.Rmd`) scripts  
* Please see the `rmarkdown.Rmd` or `rmarkdown.nb.html` files for details on rmarkdown scripts  

## Data structures (1): scalars and vectors  

* Vectors are simple lists with each item being of the same type.
* Scalars are really just vectors with a single item.
* Some useful vectors are 'built' into R:
* (place the cursor anywhere in a line of code and press Ctrl-Enter to execute
* the code / Cmd-Enter on a Mac)
* Results / output will appear in the 'Console' window below the script window.



```{r}
letters
```

```{r}
LETTERS
```

```{r}
month.abb
```

```{r}
month.name
```


Vectors can be generated using the c() function:
```{r}
c(14, 6, 2016, 12)
```

## Functions  
***Functions are used to manipulate data*** 

### Standard arithmetic  
R includes all the standard arithmetic function symbols: + - * / ^  
```{r}
3 + 4
```

```{r}
300 / 100
```

```{r}
10 - 9
```

```{r}
2 ^ 4
```

Modulo:  
```{r}
12 %% 5
```

### Names functions  
Named R functions are very similar to worksheet functions in Excel:  

1. Each function has a different name  
2. Function names are always followed by brackets  
3. Information typed between the brackets are called "arguments"  
4. Two or more arguments are separated by commas  

To see a list of all (>1200) built-in functions that come with R (aka "base R") type:  
```{r}
help(package="base")
```

So `help()` is a first example of a function. ' package="base" ' is the argument here.

### Some examples of frequently used functions  
(note how in some the examples we use vectors as arguments)  

```{r}
c(3, 4, 10)
```

```{r}
sum(c(3, 4, 10))
```

```{r}
sqrt(121)
```

```{r}
mean(c(3, 4, 10))
median(c(3, 4, 10))
max(c(3, 4, 10))
min(c(3, 4, 10))
```

## Variables  
Usually data are stored in named VARIABLES.  

* Variables are defined with the '<-' assignment operator  
* Variable names can be (almost) any length  
* Variables CANNOT contain spaces  
* Variables CANNOT contain dashes/hyphens    
* Variables CANNOT start with a number (but numbers can appear elsewhere in the variable name)  

### Two scalars:
```{r}
my.first.variable <- "Hello world!"
my.second.variable <- 2019
```

***Note that variables are listed in the 'Environment' tab in the top right RStudio panel.***  

Use the `print()` or `cat()` function to display the contents of a variable.  
Output from `print()` is preceded by row numbers in square brackets; with `cat()` the row numbers are not printed:

Using `print()`:  
```{r}
print(my.first.variable)
print(my.second.variable)
```

Using `cat()`:
```{r}
cat(my.first.variable)
```


```{r}
cat(my.second.variable)
```

Vectors can be generated using the `c()` function:
```{r}
v1 <- c(14, 6, 2016)
```

Typing just the variable is equivalent to `print()`:
```{r}
v1
```

The `View()` function opens a new tab - mostly used to inspect large tables.  
The tab will include several useful ways of inspecting and filtering a table.
```{r}
View(mtcars)
```

## Playing with vectors  
**Note**: The objective section is just to illustrate the usefulness of functions and programming languages in general. You do ***not*** need to remember any of the specifics!  

### Playing with text vectors  
```{r}
v2 <- c("Hello", "world", "!")
v2
```

Vector items can be 'glued' together into a single piece of text using the `paste()` function:
```{r}
paste(v2, collapse=" ")
```

Modifying capitalization:  
```{r}
toupper(v2)
tolower(v2)
```

### Playing with number vectors  
It's easy to generate vectors with simple series of numbers:

```{r}
v3 <- 1:12
v3
```

The seq command can generate more complicated number vectors:
```{r}
seq(from=1, to=22, by=3)
```

```{r}
vv <- seq(from=1, to=10, length.out=19)
vv
```

The `length()` function returns the number of items in a vector:  
```{r}
length(vv)
```

### Plotting vectors  
Plotting vectors is very straightforward:  
```{r}
plot(1:50)
```

```{r}
t <- (1:100) / 10
plot(sqrt(t) * cos(t), sqrt(t) * sin(t), type="l", axes=FALSE, col="blue")
```


### Vector items can have names  
The `names()` function is used both to define and to return the names of the items of a data object.
```{r}
v3
month.abb
```

Define names:  
```{r}
names(v3) <- toupper(month.abb)
v3
```

Return names:
```{r}
names(v3)
```

_**In R it is very easy to manipulate all items in a vector with simple code:**_  
```{r}
v3 * 10
```

