---
title: "Experimental Design and Data Analysis 2021/22"
subtitle: "(03) R Data Types"
author: "Axel Nohturfft"
date: "`r Sys.Date()`"
output: 
  html_notebook:
    toc: TRUE
    toc_float: TRUE
    toc_depth: 3
    code_folding: "show"
---

## Topics covered in this script  

**Data types in R**:  

The most important data TYPES in R are:  

1. numeric  
2. text (aka 'character' or 'strings')  
3. Boolean (aka 'logical')  

*(numeric data are split further into sub-categories but we don't to worry
about that here)*  

```{css echo=FALSE}
h1 {background: brown;color: white;padding-left: 7px;}
h2 {background: grey;color: white;padding-left: 7px;}
h3 {color: brown;}
code {color: chocolate;}
```

```{r echo=FALSE}
# Clear variables in memory
rm(list = ls(all=TRUE))
```


## Numerical data  
```{r}
v3
is.numeric(v3)
class(3.5)
# typeof(v3) # integer = whole numbers
# typeof(3.5) # "double"
```

## Text data  
Text format in R is called 'character':  
```{r}
v2
class(v2)
# is.character(v2)
# is.numeric(v2)
# typeof(v2)
```

## Boolean/logical data  
Boolean/logical values are often generated through 'equal' or 'greater/lesser than' (<==>) operations:  

```{r}
10 > 2
2 > 10
3 == 3.0
4 == 5
```
The following generates a vector of boolean values:
```{r}
1:10         # generates a vector of numbers from 1 to 10
1:10 >= 5    # generates a vector of boolean values
```

Store in a vector:  
```{r}
v5 <- 1:10 >= 5
v5
```

To check and illustrate let's add names for the vector items:
```{r}
names(v5) <- paste("No.", 1:10, sep="")
v5
```

Boolean values in R are called "logical":  
```{r}
class(v5)
is.logical(v5)
# is.logical(letters)
# typeof(letters)
# is.numeric(1:10)
# is.numeric(1.2)
```