---
title: "Experimental Design and Data Analysis 2021/22"
subtitle: "(04) R Data Structures"
author: "Axel Nohturfft"
date: "`r Sys.Date()`"
output: 
  html_notebook:
    toc: TRUE
    toc_float: TRUE
    toc_depth: 3
    code_folding: "show"
---

## Topics covered in this script  

* Scalars and vectors  
* Matrices  
* Calculations with vectors + matrices  
* Data frames  
* Merging ("joining") data frames  


```{css echo=FALSE}
h1 {background: brown;color: white;padding-left: 7px;}
h2 {background: grey;color: white;padding-left: 7px;}
h3 {color: brown;}
code {color: chocolate;}
```

```{r echo=FALSE}
# Clear variables in memory
rm(list = ls(all=TRUE))
```

## Data structures (1): scalars and vectors  

* Vectors are simple lists with each item being of the same type.
* Scalars are really just vectors with a single item.
* Some useful vectors are 'built' into R:
* (place the cursor anywhere in a line of code and press Ctrl-Enter to execute
* the code / Cmd-Enter on a Mac)
* Results / output will appear in the 'Console' window below the script window.

The subject was covered in `r_code_basics.Rmd`

### Vector items can have names  
The `names()` function is used both to define and to return the names of a data object.  

Let's generate a vector consisting of the numbers from 1 to 12:  
```{r}
v3 <- 1:12
v3
```

Remembers that the abbreviated months are stored in a built-in R variable called `month.abb`:
```{r}
month.abb
```

Now we set the names of the number vector to the months of the year:
```{r}
names(v3) <- toupper(month.abb)
v3
```

Return the names of the vector items:  
```{r}
names(v3)
```

_**In R it is very easy to manipulate all items in a vector with simple code:**_  
```{r}
v3 * 10
```


## Data structures (2): Matrices  

Matrices are 2-dimensional arrays/tables (think of Excel cells). In an R matrix each item (= "cell") has to be of the ***same data type***:

Let's take the numbers from 1 to 24:  
```{r}
# generates a vector
1:24
```

Convert the vector into a matrix with 6 columns  
```{r}
mx1 <- matrix(1:24, ncol=6)
mx1
```

```{r}
# Fill matrix by column:
# mx2 <- matrix(LETTERS[1:24], nrow=6, byrow=FALSE)
# mx2
```

```{r}
# Fill matrix by row:
# mx3 <- matrix(LETTERS[1:24], nrow=6, byrow=TRUE)
# mx3
```

## Calculations with vectors + matrices  

As we have already seen, one of the great things about vectors and matrices in R is that they can be modified with simple code:

```{r}
# v3
v3 * 100
```

```{r}
# mx1
mx1 + 1000
```

```{r}
# Transpose a matrix with 't()' function:
# mx1
# t(mx1)
```

## Data structures (3): data frames  

Data frames are the most common container used to store data in R.  
Each column can be a different data type.  
Each column in a data frame is in fact a vector, which means that data types cannot be mixed within a column.  

Generating a data frame 'by hand'. Note how we assign a vector to each column by name:  
```{r}
df1 <- data.frame(Name=c("Jane", "Jack", "Jaim"),
                  Age=c(5, 12, 30),
                  Female=c(TRUE, FALSE, FALSE))
df1
```

To show what data type is found in each column of the data frame use this code:
```{r}
sapply(df1, class)
```

Very large data frames are best viewed in a separate tab using the `View()` function:  
```{r}
# View(df1)
```


### Merging ("joining") data frames  

We often need to combine 2 data frames. This can be done with the `merge()` function.  

The data frames to be merged must have at least one column in common. In the example below, both the df2a and df2b data frames share the "Gene" column. The values in the shared column do not have to be in the same order.  
```{r}
df2a <- data.frame(
        Gene = c("ABCG1", "IRF1", "STAT3", "NEU1"),
        Gene.ID = c(9619, 3659, 6774, 4758)
)
df2a
```


```{r}
df2b <- data.frame(
        Gene = c("ABCG1", "IRF1", "STAT3", "NEU1"),
        Gene.Name = c("ATP binding cassette subfamily G member 1",
                      "interferon regulatory factor 1",
                      "signal transducer and activator of transcription 3",
                      "neuraminidase 1"
                      )
)
```


```{r}
df3 <- merge(x=df2a, y=df2b, by="Gene")
df3
```