---
title: "Experimental Design and Data Analysis 2021/22"
subtitle: "(05) Subsetting vectors and data frames"
author: "Axel Nohturfft"
date: "`r Sys.Date()`"
output: 
  html_notebook:
    toc: TRUE
    toc_float: TRUE
    toc_depth: 3
    code_folding: "show"
---

## Topics covered in this script  

* Subsetting vectors  
* Subsetting data frames  
* Subsetting by index  
* Subsetting using names  
* Subsetting using Boolean values  


```{css echo = FALSE}
h1 {background: brown;color: white;padding-left: 7px;}
h2 {background: grey;color: white;padding-left: 7px;}
h3 {color: brown;}
code {color: chocolate;}
```

```{r echo=FALSE}
# Clear variables in memory
rm(list = ls(all=TRUE))
```




## Subsetting (1): Subsetting vectors  

### Subsetting vectors by index  

You can select items from within a vector using square brackets + indices ...  

```{r}
# LETTERS is a built-in variable that stored the uppercase letters of the alphabet:
LETTERS
```


Subsetting with a single index number...  
```{r}
LETTERS[11]
```

Subsetting with a vector of index numbers ...  
```{r}
# The c() function generates a vector:
c(8,5,12,12,15)
```

```{r}
h <- LETTERS[c(8,5,12,12,15)]
h
```

Remember how to 'glue' vector items together?
```{r}
paste(h, collapse="")
```

### Subsetting vectors using names  
You can also refer to vector items by name, if defined ...
```{r}
v3 <- 1:12
names(v3) <- toupper(month.abb)
v3
```

```{r}
# v3[11]
v3["NOV"]
# v3["NOV"] * 10000
```

Using a vector of names:
```{r}
# c("JAN", "MAR")
# v3[c(1,3)]
v3[c("JAN", "MAR")]
```

### Subsetting vectors using Boolean (TRUE/FALSE) values  
```{r}
v4 <- 1:3
v4
```

```{r}
c(TRUE, FALSE, TRUE)
v4[c(TRUE, FALSE, TRUE)]
```

It can be easier to save a vector in a variable first, then use the variable for subsetting:
```{r}
v4
v4 > 1
x <- v4 > 1
x
v4[x]
```


## Subsetting data frames  

```{r}
df1 <- data.frame(Name=c("Jane", "Jack", "Jaim"),
                  Age=c(5, 12, 30),
                  Female=c(TRUE, FALSE, FALSE))
df1
```


### Subsetting data frames by indices  

Similar to vectors, extract data from data frames using square brackets;  
rows and columns are separated by a comma:  

* Rows: number before comma  
* Columns: number after comma  

Equivalent to "A1" (column A, row 1) in Excel:  
```{r}
df1[1,1]
```

Equivalent to "B3" (column B, row 3) in Excel:  
```{r}
df1[3,2]
```

Leaving the space before the comma blank will return entire columns (all rows):  
```{r}
df1[,1]
```

Accordingly, leaving the space AFTER the comma blank will return entire rows:
```{r}
df1[1,]
```


As with vectors, extract several rows/columns using number/index vectors:

* Before comma: `c(2,3)` = rows 2 and 3  
* After comma: `c(1,3)` = columns 1 and 3  

```{r}
df1[c(2,3), c(1,3)]
```


### Subsetting data frames using Boolean vectors  
As with vectors, we can also subset with Boolean/logical values:  

* Before comma: **`c(FALSE, TRUE, TRUE)`** =  
  - Row 1 = FALSE (don't include)  
  - Row 2 = TRUE (***do*** include)  
  - Row 3 = TRUE (***do*** include)  
* After comma: **`c(TRUE, FALSE, TRUE)`** =  
  - Column 1 = TRUE (***do*** include)  
  - Column 2 = FALSE (don't include)  
  - Column 3 = TRUE (***do*** include)  
  
```{r}
df1[c(FALSE, TRUE, TRUE), c(TRUE, FALSE, TRUE)]
```


### Extracting a single column from a data frame by name  

For data frames the `names()` command refers to column headers:
```{r}
names(df1)
```


```{r}
names(df1) <- c("Nom", "Age", "Femelle")
df1
```

We often use the dollar sign to refer to entire column of a data frame (returns a vector):
```{r}
df1$Nom
class(df1$Nom)
```


## Subsetting data frames using row and column names  

Data frames (and matrices) can have row names as well:
```{r}
row.names(df1) <- paste0("Row", seq_len(nrow(df1)))
df1
```

We can use both row names and column names to subset data frames:  
*(NOTE that text always needs to be defined with quotation marks)*  

Select 2 columns (all rows):
```{r}
df1[, c("Nom", "Femelle")]
```

Select rows (all columns):
```{r}
df1[c("Row1", "Row3"), ]
```

Select columns and rows:
```{r}
df1[c("Row1", "Row3"), c("Nom", "Femelle")]
```

Sometimes code becomes easier to read if index vectors are first stored in a variable; then use the variable to subset the data frame:
```{r}
a <- c(2,3)
b <- c(TRUE, FALSE, TRUE)
c <- c("Nom", "Age")
```

```{r}
df1[a, b]
```

```{r}
df1[b, c]
```

